---
title: "R hacks"
author: "Katerina Krizova"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: 
  pdf_document: 
    toc: yes
    toc_depth: 3
---

* frequently used commands and hardly gained experience 
* R, R Markdown

# *WHAT NEEDS TO BE COVERED*

## apply, lapply, vapply aso

# *INITIAL STEPS*

## PACKAGES

* tidyverse : 

```{r}
# install.packages("tidyverse")
require(tidyverse)
```

## DATA

* **starwars**
* **simple_table_4read.csv**
* **simple_table_2write.txt**

During this demonstration, We’ll use the *Starwars* characters dataset. 
The data, part of the dplyr package, come from [SWAPI](https://swapi.dev/), the Star Wars API. Since the data are integrated into *dplyr*, we don’t need to load the data, but you may still want to find information about the dataset. The codebood for the Starwars dataset is integrated into the dplyr documentation. To view the Starwars codebook, first load the library(dplyr) package. Then, via the Help pane, search starwars. Alternatively, in the Console pane, type ?starwars.

Note: dplyr is part of tidyverse which we loaded with the command: library(tidyverse)

```{r}
starwars
```

## SET PATHS

```{r}
path_in <- "D:/VURV/R_hacks/data/"
path_out <- "D:/VURV/R_hacks/results/"
```

# *R HACKS*

# READ AND WRITE

* read.table
* read.csv : sep is always ","

* write.table : usually set sep as ";"
* write.csv

## READ

header = T : 1st row is header

sep = "," OR ";" OR " "

read.table and read.csv is both fine

> absolute path

```{r}
df <- read.table("D:/VURV/R_hacks/data/simple_table_4read.csv", header = TRUE, sep = ";")
df <- read.csv("D:/VURV/R_hacks/data/simple_table_4read.csv", header = TRUE, sep = ";")
```

> relative paths

```{r}
df <- read.table(paste0(path_in, "simple_table_4read.csv"), header = TRUE, sep = ";")
df <- read.csv(paste0(path_in, "simple_table_4read.csv"), header = TRUE, sep = ";")
```

## WRITE

quote = F : string will be not written in quotes

row.names = F : no primary ID will be added

col.names = F : header won't be written

> absolute path
```{r}
write.table(df,"D:/VURV/R_hacks/results/simple_table_2write.txt", 
            sep = ";", quote = F, row.names = F) 
```

> relative paths

```{r}
write.table(df, paste0(path_out, "simple_table_2write.txt"), 
            sep = ";", quote = F, row.names = F) 
```

# FOLDERS, FILES AND LISTS

## create folder in direcory

```{r warning=F}
dir.create ("D:/VURV/R_hacks/results/new_folder_abs") # absolute path
dir.create (paste0(path_out, "new_folder_rel")) # relative path
```

## load all files from directory 

multiple files can be loaded from directory based on specific rule
mostly file type or name of the file

## names of files without extension

```{r}
library(tools)
file_path_sans_ext(list.files(pattern = "*.csv"))
```


## read files from direcory and combine them into 1 file

## split df based on a factor level

```{r eval=F}
sw_split <- split(starwars, starwars$homeworld) # creates list

lapply(names(sw_split), function(df) write.csv(sw_split[[df]], file=paste0(path_out, "new_folder_abs/", df, ".csv"), row.names = F)) 
```


## remove df from environment

remove single file or list of files with specific name pattern

```{r eval=F}
rm(df_round) # created later
rm(list=ls(pattern="sw_")) # created later
```


# NUMERICS

## round all

rounds all numeric values in DF

```{r}
df_round <- df %>% 
  mutate(across(where(is.numeric), ~ round(., digits = 1)))
```

## date format

[Date Formats in R](https://www.r-bloggers.com/2013/08/date-formats-in-r/)

```{r}
df$sow_dat <- toString(df$sow_dat) # convert date into date format
df$sow_dat <- as.Date(df$sow_dat, "%Y%m%d") # convert date into date format

# doesn't work so far
# df$month <- format(df$sow_dat, "%m") # get months as numeric     
# df$year <- format(df$sow_dat,format="%y") # get years
```

# STRINGS

## starts with 

```{r eval=F}
sw_string <- starwars %>% 
  filter(startsWith(homeworld, "Tat"))
```

## remove last X characters

str_sub(species,**1**,nchar(species)**-2**))
1 -> new string strats with 1st character
-2 -> cut last 2 characters

```{r}
sw_rem <- starwars %>% 
  mutate(spec_short = str_sub(species,1,nchar(species)-2))
```


# SALT & PEPPER

```{r}
summary(starwars$species[starwars$species == "Human"])
summary(starwars$height[starwars$height > 100])
summary(starwars$height) # to compare with selected part of the table
```


# DATA ANALYSIS

## delete rows with NA values

```{r eval=F}
na.omit()
```

## case_when

creates a new column based on info from other columns
```{r}
sw_casewhen <- starwars %>% 
  mutate(power = case_when(sex == "female" ~ "female power", 
                           sex %in% c("male","none") ~ "too bad")) 
```

## if else

```{r}
#quest <- "nam_hom" # names and homeworld
quest <- "body" # name, heights, mass
#quest <- "color" # name, hair_color, skin_color, eye_color 
#quest <- "nope" # name, hair_color, skin_color, eye_color


if(quest == "nam_hom"){
     sw_quest <- starwars %>% 
    select(name, homeworld)
    print("Da wo ich herkomm")
  } else if(quest == "body"){
    sw_quest <- starwars %>% 
    select(name, height, mass)
    print("No one will jugdge you")
  } else if(quest == "color"){
    sw_quest <- starwars %>% 
    select(name, hair_color, skin_color, eye_color)
    print("You are all beautiful")
  } else{
    print("It's a trap!")
  }
```


# GGPLOT

## variables

## lab names

## lab scales

## legend name

## legend position

## facet

### facet_grid

### facet_wrap

## ggsave

```{r eval=FALSE}
ggsave("D:/VURV/R_hacks/results/plot.png", device = "png", width = 10, height = 6, dpi = 300)
ggsave(paste0(path_out, "plot.png"), device = "png", width = 10, height = 6, dpi = 300)
ggsave(paste0(path_out, "plot.png"), device = "png", width = 10, height = 6, dpi = 300) # add variable to the name
```


# MARKDOWN

## document header

title: "Habits"
author: John Doe
date: March 22, 2005
output:
  pdf_document
    toc: true
    toc_depth: 2
    number_sections: true
    fig_width: 7
    fig_height: 6
    fig_caption: true
  fontsize: 11pt
  geometry: margin=1in


## r chunk options

* \`\`\` \{r\} \`\`\`
* r include = F : prevents from including the chunk in knitted file
* r echo = F : code not in file, results yes
* r eval = F : code is in the file, but the chunk does not run
* r message = F : does not show the message (e.i. "shp uploaded")
* r warning = F : does not show the warning (e.i. "folder already exists")

## headers

\# Title 1

\#\# Title 2

\#\#\# Title 3

\*text\*  *italics*

\*\*text\*\* **bold**

